GravExplain: Demonstrating continuous gravitational wave searches using sound and a table-top interferometer

### 2020-02-14 paper/paper_main.tex
PhB final submission

% Note that AJP uses the same style as Phys. Rev. B (prb).
% The AIP Style Manual\cite{AIPstylemanual} is an indispensable reference on good physics writing, covering everything from planning and organization to standard spellings and abbreviations.
% Most important of all, please familiarize yourself with the AJP Statement of Editorial Policy,\cite{editorsite} which describes the types of manuscripts that AJP publishes and the audience for which AJP authors are expected to write.
% We look forward to receiving your submission to AJP.
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{float}
\usepackage{mathtools}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
% \usepackage{siunitx}

@@

% cover page
\section*{Project contributions}
% contributions clarify what’s to be assessed
James W. Gardner:
\begin{itemize}
\item all paper sections other than the introduction, i.e.\ from Section~\ref{sec:ifo} onwards, and all figures in the paper
\item addition of the injection speaker, design of photodiode circuit from standard examples, assembly of circuit on breadboard, and set-up of Raspberry Pi to measure output
\item all of the source code at \url{https://github.com/daccordeon/gravexplain}, that reads the data from the Raspberry Pi, analyses it, and creates the figures used in the paper
\item talks presented were based on the paper and figures therein
\end{itemize}

Hannah Middleton:
\begin{itemize}
\item introduction of paper, i.e.\ Section~\ref{sec:introduction}
\item design and assembly of the interferometer
\end{itemize}

Andrew Melatos, Robin Evans, William Moran: vital advice and comments throughout the project

\subsection*{Statement of achievements}
% one paragraph recount of what’s been done for two months, like the abstract
The project spanned three weeks in December, four weeks in January, and two weeks in February. The first two weeks were spent writing the initial scripts and achieving the Viterbi results with the webcam set-up. The third week was spent setting up the photodiode circuit and the Raspberry Pi. All of January was spent progressing through the optical microphone experiment, and was largely driven by suggestions made at the signal processing meetings with the electrical engineers. By the end of the third week of January we had a working optical microphone but low speech intelligibility. The fourth week of January was spent on attempting methods of speech enhancement. Finally, the two weeks in February were spent writing and drafting the paper/report, and giving presentations.

@@

\title{GravExplain: Continuous gravitational wave searches in a table-top experiment}

\author{James W. Gardner}
\email{u6069809@anu.edu.au}
\affiliation{Research School of Physics, Australian National University, 60 Mills Rd, Acton ACT 2601 Australia}

@@

% AJP requires an abstract for all regular article submissions.
\begin{abstract}
Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy interferometric data. We demonstrate some of these techniques by observing sound with a table-top Michelson interferometer. In particular, we use the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
We also demonstrate how the inclusion of a photodiode and Raspberry Pi allows the interferometer to act as an optical microphone. That is, we are able to play back simple audio recordings from the photodiode data. However, more complex recordings, of speech, are unintelligible when played back, despite preliminary efforts towards speech enhancement.

@@

% general gw intro
In 2015 gravitational waves were observed for the first time from the merger of two black holes in a binary system.~\cite{GW150914} 
The observation, made by the Laser Interferometer Gravitational-wave Observatory~\citep[LIGO]{AdvancedLIGO:2015}, marked a breakthrough in modern astrophysics and provides a new means to observe the universe. 
Since 2015, the LIGO and Virgo~\cite{AdvancedVirgo:2015} observatories have made numerous detections of binary black hole~\cite{GW151226,GW170104,GW170814} and binary neutron star~\cite{GW170817,GW170817multi,GW190425} mergers. 
In recent years there has been an increased public interest in gravitational-wave science. 
Many gravitational wave research groups around the world have produced demonstrations and activities to aid in explaining this topic to a general audience.
Activities range from online data analysis tutorials,~\cite{GWOSC:online,LOSC:2015} hands-on demonstrations, phone apps,~\cite{LaserLabs:online,SciVR:online} and online games,~\cite{BlackHoleHunter:online} to exhibitions,\cite{L2URSSE} artistic interpretation, and musical performance~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online}.

% what are gws and how do we detect them...
Gravitational waves are a prediction of Einstein's theory of general relativity. 
They are disturbances in spacetime caused by the acceleration of massive objects. 
The effect of gravitational waves is a change in distances; a ``stretch and squeeze'' effect. 
Observatories such as LIGO, Virgo, and KAGRA~\cite{KAGRA:2013} are laser interferometers; they use the interference of laser light to measure changes in distance. 
Gravitational-wave observatories are extremely complex devices, however they are based on the Michelson interferometer. 
Table-top Michelson interferometers are commonly used by gravitational-wave research groups to demonstrate the science of gravitational-wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,ExhibitIFO,LIGOIFOGlue,LIGOIFOMagnets} and they are often used in undergraduate lab experiments~\cite{UgoliniEtAl:2019}.

% continuous gws
To date the network of gravitational-wave observatories have observed short-duration transient signals~\cite{GWTC-1:2018,GWOSC:online}. 
These observatories are also searching for continuous gravitational waves; persistent periodic signals at near-monochromatic frequencis. 
One of the prime targets of continuous wave searches are rotating neutron stars in low mass X-ray binaries (LMXBs).
LMXBs are binary systms containing a compact object (such as a neturon star or a black hole) in orbit with a low mass stellar companion~\cite{xraybinaries:1997}. 
The LMXB Scorpius X-1 is an LMXB which is particularly bright in X-rays. 
It is a prime target for continuous wave searches from LMXBs and numerous searches have been performed~\cite{2019PhRvD.100l2002A,RadiometerO1O2:2019,SearchCrossCorrO1:2017}.

% sound analogy
Here we demonstrate the use of a Michelson interferometer as a tool to explain the search for continuous waves to a general audience.
The analogy of listening to the universe is often used to explain gravitational-wave observations; transient gravitational-wave signals have been converted into audio.
Continuous waves may emitt at constant frequencies, or they may wander slowly in frequency over time. 
This can be demonstrated with a table-top interferometer by playing audio of constant or slowly changing frequency tones to produce a simulation of a continuous wave signal. 
The analysis techniques used in this work are similar to those used to search for continuous gravitational waves in LIGO and Virgo data, particularly for the wandering frequency demonstration~\cite{2019PhRvD.100l2002A,ScoX1ViterbiO1:2017,SuvorovaEtAl:2017,SuvorovaEtAl:2017}.

In this work we detail our table-top intererometer design in Section~\ref{sec:ifo}, in Section~\ref{sec:single_tone} we describe observing a single tone signal, Section~\ref{sec:viterbi_wandering} describes observing a wandering frequency signal, and Section~\ref{sec:optical_microphone} extends the set-up to be able to capture complex audio, such as music and speech. We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}.

@@

Detectable gravitational waves are caused by some of the most violent events in the universe. Like two massive black-holes merging with each other\cite{GW150914}. Even so, the effects they produce are on the order of one part in $10^{21}$ for our best detectors\cite{GW150914}. For a table-top experiment these effects would be far too small, so we demonstrate observing a far louder source, sound waves. Although the analogue to gravitational waves is tenuous, we can still demonstrate some of the continuous wave analysis satisfactorily. The other benefit of observing sound is that we can use existing signal processing techniques designed for audio signals.

\subsection{Michelson interferometry}
% hardware: path difference, virtual thin film interference

A Michelson interferometer is a well-known optics configuration used in many undergraduate and research labs, including in the detectors for gravitational waves. A laser beam of coherent (monochromatic and equal phase) light is split evenly by a 50/50 beamsplitter into two arms at right-angles to each other, travelling down and back from a mirror at the end of each arm. The beams recombine and interfere at the beamsplitter before hitting a screen to produce a pattern of concentric, circular fringes. The configuration is shown in Figure~\ref{fig:ifo_schematic_webcam}. In the detectors for gravitational waves, these arms are kilometres long and many other complexities (such as power recycling and quantum squeezing)\cite{GW150914} are necessary to observe a signal.

\begin{figure}
	\includegraphics[width=0.8\textwidth]{figures/ifo_schematic_webcam.pdf}
	\caption{Michelson interferometer schematic, interference pattern is recorded by a webcam}
	\label{fig:ifo_schematic_webcam}
\end{figure}
% Note that we do not include the common compensating plate that equalises the amount of time spent in glass for each beam. We did not find it necessary to include it to be able to demonstrate the results desired.

The interference pattern, as shown in Figure~\ref{fig:interference_pattern}, is due to the path difference between the two arms of unequal lengths (if the lengths were equal, then only a single, bright dot would be seen). This results in a phase difference between the two beams when they meet at the screen. This path difference is perhaps better seen as the perpendicular separation between one mirror and the virtual image of the other.
For an aligned configuration and a fixed separation, the phase difference at a point only depends on the angle from the centre-line of the beam, and so the pattern is circularly symmetric.

\begin{figure}
	\includegraphics[width=\textwidth]{figures/webcam_still0.pdf}
	\caption{Interference pattern from Michelson interferometer as viewed by webcam from off-centre (hence why the pattern appears non-circular), red dot indicates point where intensity time series was measured from. For scale, the central fringe is approximately 5mm across}
	\label{fig:interference_pattern}
\end{figure}

@@

Continuous gravitational wave searches are, by the nature of the astrophysical sources that they look for, always searching for periodic, or even sinusoidal, signals. Often the frequency of this periodicity can change, as addressed later, but sometimes it is (near) constant. As such, the audio analogue is to observe a single tone (a sinusoidal note of one frequency).

\subsection{Method}

We use a 0.5W speaker stuck to the back of one of the mirrors as a source for our demonstration, as shown in Figure~\ref{fig:ifo_schematic_webcam}. This speaker was part of a commercial 3.5mm-jack, USB powered pair of speakers and is driven by a PC. Previously, we used a function generator to accomplish the same signal.
As the speaker deforms to play sound, the mirror moves with a change in separation, $\delta d$, as some coupling function of the amplitude and frequency of the sound. The motion of the fringes follows the motion of the mirror and therefore the speaker at the same frequency.
If the fringe motion is small, then changes in the intensity at some point on the screen follow the frequency of the injected (played) audio but with amplitude given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
Large fringe motion means that multiple fringes can pass over the measured point in a single speaker deformation, such over-counting would artificially raise the measured frequency. As such, any motion of the fringes, and so of the speaker, must be kept small by playing sound softly through the speaker.

@@

We measure the intensity at a point on the screen with a commercial USB webcam placed as to view the screen (Figure~\ref{fig:ifo_schematic_webcam}), sampling at 30Hz (or 30 fps, frames-per-second). This limits the spectral content of signals we can observe to be under the Nyquist frequency of 15Hz.
See later Section~\ref{sec:photodiode} where we raise this limit into to audible frequencies (20 Hz to 20kHz) with a photodiode.
While playing a tone under 15Hz through the speaker for a minute, we recorded the interference pattern with the webcam. Then, at an off-centre pixel as seen in Figure~\ref{fig:interference_pattern}, we take the green-channel of the colour (RGB) video as approximate to the total intensity, as the laser is monochromatic green light. We can play this time series of intensity over time as an audio recording (we used the \textbf{scipy.io.wavfile.write}\cite{scipy} function in Python\cite{python}) and the tone is clearly heard.
Taking the Fourier transform of this signal shows the injected tone, as seen in Figure~\ref{fig:webcam_spectrum} where we record the highest peak Fourier amplitude at $(2.099\pm 0.008)\mathrm{Hz}$ for an injected tone of 2.09Hz (a result within two standard deviations).
Therefore, we’re able to observe a single tone quite confidently.

\begin{figure}
	\includegraphics[width=\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{Fourier amplitudes of intensity time series of interference pattern as captured by a webcam for an injected tone of 2.09Hz, the highest peak is measured at $(2.099\pm 0.008) \mathrm{Hz}$ for an injected tone of 2.09Hz. In fact, the actual highest peak is at 0Hz, but that corresponds only to a non-zero average value, so we don’t display or consider it.}
	\label{fig:webcam_spectrum}
\end{figure}

Continuous gravitational wave searches often look for signals that wander (change) in frequency slowly over time\cite{2019PhRvD.100l2002A}. This means that finding the signal isn’t as simple as searching for a peak in the frequency content for the full observing run, especially when the signal-to-noise ratio is particularly low.

One method (among many) is to split the observed time series into bins of time, compute the Fourier transform for each time interval, and then create a grid (known as a spectrogram) where each cell is the Fourier amplitude of a particular frequency at a particular time.
The method then treats the problem as one of finding the best path through a weighted graph, where here the weights are given by the Fourier amplitudes.
The method takes in a parameter of the allowed frequency wandering, some rate of frequency change over time, usually informed by a physical model. Here, this parameter is set to 0.11 Hz per second to make it simply one frequency bin per every time bin, which will prove poorly adjusted to the signal.

The Viterbi algorithm is a path-finding algorithm that solves this problem.
Abstractly speaking, the algorithm is given a weighted graph (e.g.\ the spectrogram grid), a sensible sequence of subgraphs (e.g.\ the columns holding spectrum at each time), and restrictions on connectivity (e.g.\ the allowed amount frequency wander). Then, at each iteration, it will find the best path to each node in the next subgraph, all the way from the first subgraph. At the end, it selects the overall best path from the first to the last subgraph (here, from the start to end time), this overall best path is called the Viterbi path. See Appendix~\ref{app:viterbi} for the full details of the algorithm in this particular implementation.

We demonstrate this method of using the Viterbi algorithm to find a signal, by injecting a wandering frequency signal into the speaker and recording the output via the webcam.
Note that when creating a wandering frequency signal, the phase change must be integrated up over time, as detailed in Appendix~\ref{app:phase_gotcha}.

Shown in Figure~\ref{fig:viterbi_overlay} is the resultant spectrogram of the observed signal, overlaid with the injected signal and the recovered Viterbi path.
To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment. Quantifying this, the Viterbi path stays within one frequency bin (approximately 0.3Hz) of the injected tone for 94\% of the run.
Note that the signal initially moves faster than the chosen frequency wander parameter, causing a discrepancy between the injected and recovered signals. To fix this, the parameter should be doubled, although this would then cause the recovered path to stray more often. There also appears to be an anomaly at 150s, likely from a disturbance (such as someone walking) around the interferometer.

Empirically, decreasing the signal-to-noise ratio closer to unity also causes the Viterbi algorithm to more easily stray from the signal to reach nearby fluctuations. However, the only correlation through the grid is the signal, so we always expect the Viterbi path to lie near to the signal no matter how high the noise is.

\begin{figure}
	\includegraphics[width=\textwidth]{figures/expt_overlay_2_viterbi_test_webcam.pdf}
	\caption{Spectrogram of observed signal overlaid with white crosses for the Viterbi path and pink dots for the injected signal, notice the high correlation between them indicating that the signal is able to be recovered; also note the deviation at the start when the signal moves too quickly and just before the background anomaly at 150s}
	\label{fig:viterbi_overlay}
\end{figure}

@@

The natural succession to observing simple tones is to study complex audio signals, such as music and speech.
The Viterbi algorithm and tracking more broadly is not applicable to these signals. Compare a single tone that slowly, continuously changes frequency to music and speech which consist of a broad spectrum that is constantly changing.
Instead, the interest here is in being able to record and play-back these complex audio signals, rather than track them. Using the interferometer to do so makes it, in ensemble with the measuring device, an ‘optical microphone’, using light to capture sound.

Advanced signal processing is required to extract gravitational wave signals from noisy observational data. Recovering audio from an optical microphone requires related signal processing. Although far simpler, it serves as another useful demonstration.

@@

For an optical microphone to be sensitive to speech and music it needs a frequency response (a sampling rate) far higher than the 30Hz of a standard webcam. Both speech and music need at least kHz range, with speech intelligibility requiring up to 3kHz and music preferably up to and beyond 8kHz\cite{speech_intelligibility}. This can be addressed by using a high-speed webcam, or more simply, by using a photodiode as the measuring device.

A photodiode is an electrical component that acts as a regular diode (blocking any current flow in the reverse direction) when no light is incident on it, but becomes more and more conductive in the reverse direction as the intensity of incident light increases. 
% Fundamentally, a current is created in the photodiode by the photo-electric effect.
Placing a photodiode in reverse-bias over an op-amp creates a photo-detector that will produce a voltage dependent on the incident intensity. This experiment had the photodiode in the interference pattern at roughly the same spot as the webcam was monitoring previously, just off-centre in the position of the red dot in Figure~\ref{fig:interference_pattern}.

>>>>

% Note that AJP uses the same style as Phys. Rev. B (prb).
% The AIP Style Manual\cite{AIPstylemanual} is an indispensable reference on good physics writing, covering everything from planning and organization to standard spellings and abbreviations.
% Most important of all, please familiarize yourself with the AJP Statement of Editorial Policy,\cite{editorsite} which describes the types of manuscripts that AJP publishes and the audience for which AJP authors are expected to write.
% We look forward to receiving your submission to AJP.
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{float}
\usepackage{mathtools}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
% \usepackage{siunitx}
\usepackage{textcomp}

@@

% cover page
\section*{Project contributions}
% contributions clarify what’s to be assessed
James W. Gardner:
\begin{itemize}
\item all paper sections other than the introduction, i.e.\ the abstract and from Section~\ref{sec:ifo} onwards, and all figures in the paper
\item addition of the injection speaker, design of photodiode circuit from standard examples, assembly of circuit on breadboard, and set-up of Raspberry Pi to measure output
\item all of the source code at \url{https://github.com/daccordeon/gravexplain} (containing over 1000 lines of Python scripts), that reads the data from the Raspberry Pi, performs all of the analysis on it, and creates all the figures used in the paper
\item presentation of two talks based on the paper and the figures therein
\end{itemize}

Hannah Middleton:
\begin{itemize}
\item the introduction of the paper, i.e.\ Section~\ref{sec:introduction}
\item design and assembly of the interferometer
\item continuous support and advice throughout the project
\end{itemize}

Andrew Melatos, Robin Evans, William Moran: other vital advice and comments throughout the project

\newpage
\section*{Statement of achievements}
% one paragraph recount of what’s been done for two months, like the abstract
The project spanned three weeks in December, four weeks in January, and two weeks in February. The first two weeks were spent writing the initial scripts and achieving the Viterbi results with the webcam set-up. This was relatively straight forward and the greatest hurdle was understanding the algorithm.

The third week was spent setting up the photodiode circuit and the Raspberry Pi which involved constant testing and changing of the components and design of the circuit.
All of January was spent on the optical microphone. Much difficultly was had combating 50Hz mains hum in the experiment. Eventually through many discussions at the signal processing meetings, we had a working optical microphone but low speech intelligibility. The last week of January was spent on methods of speech enhancement. If the investigation had continued we had much future work planned and possible solutions in mind.

Finally, the start of February was spent writing the paper and giving presentations (one nine minute general talk and one 50 minute technical follow-up talk).

@@

\title{GravExplain: Demonstrating continuous gravitational wave searches using sound and a table-top interferometer}

\author{James W. Gardner}
\email{u6069809@anu.edu.au}
\affiliation{College of Science, Australian National University, Acton, ACT, 2601, Australia}

@@

% AJP requires an abstract for all regular article submissions.
\begin{abstract}
Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy interferometric data. We demonstrate these techniques by observing sound with a table-top Michelson interferometer. In particular, we used the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
We also describe how the inclusion of a photodiode and Raspberry Pi allowed the interferometer to act as an `optical microphone' with sufficient fidelity to play back simple audio recordings. However, more complex recordings, of speech, remain unintelligible when played back, despite preliminary efforts made towards speech enhancement.

@@

\begin{figure}
	\includegraphics[width=0.9\textwidth]{figures/ligo_pic.pdf}
	\caption{Aerial photo of the LIGO Hanford facility and its 4km long beam arms, courtesy of LIGO/Caltech/MIT}
	\label{fig:ligo_pic}
\end{figure}

% general gw intro
In 2015, gravitational waves were observed for the first time from the merger of two black holes in a binary system~\cite{GW150914}. 
The observation, made by the Laser Interferometer Gravitational wave Observatory~\citep[LIGO]{AdvancedLIGO:2015}, marked a breakthrough in modern astrophysics and revealed a new means to observe the universe. 
Since 2015, the LIGO and Virgo~\cite{AdvancedVirgo:2015} observatories have made numerous detections of binary black hole~\cite{GW151226,GW170104,GW170814} and binary neutron star~\cite{GW170817,GW170817multi,GW190425} mergers. 
In recent years, there has been increased public interest in gravitational wave science. 
Many gravitational wave research groups around the world have produced demonstrations and activities to explain the topic to a general audience.
Activities range from hands-on demonstrations, exhibitions, and online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, to phone apps~\cite{LaserLabs:online,SciVR:online}, video games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online}.

% what are gws and how do we detect them...
Gravitational waves are a prediction of Einstein's theory of general relativity. 
They are disturbances in space-time caused by the acceleration of massive objects. 
The effect of gravitational waves is a change in lengths; a `stretching and squeezing' of the very distance between objects.
Observatories such as LIGO, Virgo, and KAGRA~\cite{KAGRA:2013} are laser interferometers; they use the interference of laser light to measure changes in distance. As shown in an aerial view of LIGO Hanford in Figure~\ref{fig:ligo_pic}, they use beam arms kilometres long.
These observatories are extremely complex, but are fundamentally are based on the Michelson interferometer. 
Table-top Michelson interferometers are commonly used in undergraduate lab experiments~\cite{UgoliniEtAl:2019} and to demonstrate the science of gravitational wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,ExhibitIFO,LIGOIFOGlue,LIGOIFOMagnets}.

% continuous gws
To date, while the network of gravitational wave observatories have only observed short-duration transient signals~\cite{GWTC-1:2018,GWOSC:online}, these observatories are also searching for continuous gravitational waves; persistent periodic signals at near-monochromatic frequencies.
These continuous waves may emit at constant frequencies, or they may wander slowly in frequency over time. 
One of the prime targets of continuous wave searches are rotating neutron stars in low mass X-ray binaries (LMXBs).
LMXBs are binary systems containing a compact object (such as a neutron star or a black hole) in orbit with a low mass stellar companion~\cite{xraybinaries:1997}. 
The LMXB Scorpius X-1 is an LMXB which is particularly bright in X-rays. 
It is a prime target for continuous wave searches from LMXBs and numerous, as yet unsuccessful, searches have been performed~\cite{2019PhRvD.100l2002A,RadiometerO1O2:2019,SearchCrossCorrO1:2017}.

% sound analogy
Here, we demonstrate the use of a Michelson interferometer as a tool to explain the search for continuous waves to a general audience; the analogy of listening to the universe is often used to explain gravitational wave observations.
 %; transient gravitational wave signals have been converted into audio.
This can be demonstrated with a table-top interferometer by playing sound close enough to move the mirrors and change the interference pattern.
Using audio of slowly wandering frequency tones to produce an analogue of a continuous gravitational wave signal.
The analysis techniques used in this work are similar to those used to search for continuous gravitational waves in LIGO and Virgo data~\cite{2019PhRvD.100l2002A,ScoX1ViterbiO1:2017,SuvorovaEtAl:2017,SuvorovaEtAl:2017}, particularly for the wandering frequency demonstration.

In this work we detail our table-top interferometer design in Section~\ref{sec:ifo} and describe observing a single note from a speaker in Section~\ref{sec:single_tone}. In Section~\ref{sec:viterbi_wandering}, we demonstrate the analysis techniques for observing a wandering frequency signal.
In Section~\ref{sec:optical_microphone}, we demonstrate capturing and playing back complex audio, such as music and speech. This demonstration of an `optical microphone' serves as a more general exhibition of signal processing. We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}.

@@

Even though gravitational waves are caused by some of the most massive events in the universe, their observable effects on Earth are vanishingly small | on the order of only one part in $10^{21}$ for our best detectors~\cite{GW150914}. Clearly, the measurement of such effects is far beyond any table-top experiment. But, the fundamental principle of gravitational wave detection remains that changes in the lengths of the beam arms of an interferometer leads to a change in the interference pattern observed. By observing sound waves that vibrate the mirrors of a table-top interferometer and so change the lengths of the beam arms we have an analogue of this principle.

\subsection{Method of Michelson interferometry to observe sound}
% hardware: path difference, virtual thin film interference

In a Michelson interferometer, a coherent (monochromatic and equal phase) source of laser light is split evenly by a beamsplitter into two arms at right-angles to each other, travelling down and back from a mirror at the end of each arm. The beams recombine and interfere at the beamsplitter before hitting a screen to produce a pattern of concentric, circular fringes. The configuration used in this experiment is shown diagrammatically in Figure~\ref{fig:ifo_schematic_webcam} and as part of a photo in Figure~\ref{fig:setup_pic2}.

A sample interference pattern is shown in Figure~\ref{fig:interference_pattern}.
This pattern is due to the path difference between the two arms of unequal lengths (if the lengths were equal, then only a single, bright dot would be seen). This results in a phase difference between the two beams when they meet at the screen. This path difference is perhaps better seen as the perpendicular separation between one mirror (M1) and the virtual image (M2\textquotesingle) of the other. For an aligned configuration and a fixed separation, the phase difference at a point on the screen only depends on the angle from the centre-line of the beam, and so the pattern is circularly symmetric. When the path difference changes, the fringes stay circular but move radially in or out.

\begin{figure}[h]
	\includegraphics[width=0.8\textwidth]{figures/ifo_schematic_webcam.pdf}
	\caption{Michelson interferometer schematic with interference pattern recorded by a commercial webcam. We used a Class 2 532nm laser~\cite{ThorLabsIFO} and a commercial speaker. The beam is focused at a converging lens before the 50/50 beamsplitter and the interference pattern is made visible by a diverging lens after the beamsplitter. The arms lengths between the speakers are roughly 3cm different and the entire interferometer (without the screen) fits within the dimensions of an A4 piece of paper. We did not include a compensating plate with the beamsplitter, as is often done.}
	\label{fig:ifo_schematic_webcam}
\end{figure}

We used a 0.5W speaker stuck to the back of one of the mirrors as an audio source for our demonstration. This speaker was part of a commercial 3.5mm-jack, USB powered pair of speakers and was driven by a PC. To attach the speaker to one of the mirrors, we wedged it into part of the mirror mount and used commercial adhesive putty to keep it in place. We kept the other speaker in the pair face-down and as far away from the experiment as the cables would allow to reduce it acting as a second source.

As the speaker deforms to play sound, the mirror moves with a change in separation as some coupling function of the amplitude and frequency of the sound. The motion of the fringes follows this motion of the mirror and therefore the speaker at the same frequency.
If the fringe motion is small, then changes in the intensity at some point on the screen follow the frequency of the injected audio but with amplitude given by a transfer function that accounts for the coupling and resonance of the speaker-mirror system.
If the fringe motion was large enough for multiple fringes to pass over the measured point in a single speaker deformation, then over-counting would artificially raise the measured frequency. As such, any motion of the fringes must be kept small by playing sound softly through the speaker.

For later sections, we derive the non-linear relation between the separation of the mirrors and the intensity of the pattern at some point on the screen, this can be found in Appendix~\ref{app:intensity_derivation}.

% Note that we do not include the common compensating plate that equalises the amount of time spent in glass for each beam. We did not find it necessary to include it to be able to demonstrate the results desired.

\begin{figure}[h]
	\includegraphics[width=\textwidth]{figures/webcam_still0.pdf}
	\caption{Interference pattern from Michelson interferometer as viewed by webcam from off-centre (hence why the pattern appears non-circular), red dot indicates point where intensity time series was measured from. For scale, the central fringe is approximately 5mm across}
	\label{fig:interference_pattern}
\end{figure}

@@

As mentioned in Section~\ref{sec:introduction}, continuous gravitational wave searches look for near monochromatic signals. The audio analogue of this is a simple sinusoidal tone; a constant note.

We measured the intensity at a point on the screen with a commercial USB webcam placed as to view the screen as in Figure~\ref{fig:ifo_schematic_webcam}. This webcam samples at 30Hz, which limits the spectral content of signals we can observe to below 15Hz, the Nyquist frequency.
% See later Section~\ref{sec:photodiode} where we raise this limit into audible frequencies (20 Hz to 20kHz) with a photodiode.
While playing a tone under 15Hz through the speaker for a minute, we recorded the interference pattern with the webcam. Then, at an off-centre pixel (shown as a red dot in Figure~\ref{fig:interference_pattern}), we took the green-channel of the video as approximate to the total intensity, as the laser is monochromatic green light (at 532nm).

We played this time series of intensity over time as an audio recording using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. The two tones played together sound similar and don’t produce any noticeable beating.
To quantify the precision, we recorded a signal from an injected tone of 2.09Hz. From the Fourier transform shown in Figure~\ref{fig:webcam_spectrum}, we measured a peak amplitude at $(2.099\pm 0.008)\mathrm{Hz}$, a result within two standard deviations. This showed that the experiment observes a single tone confidently.

\begin{figure}[h]
	\includegraphics[width=\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{Fourier amplitudes of intensity time series of interference pattern as captured by a webcam for an injected tone of 2.09Hz, the amplitude 17.5 peak was measured at $(2.099\pm 0.008) \mathrm{Hz}$. The actual maximum is achieved at 0Hz, but corresponds to the signal having a non-zero average value, so we don’t display or consider it.}
	\label{fig:webcam_spectrum}
\end{figure}

As also mentioned in Section~\ref{sec:introduction}, continuous wave searches often look for signals that wander slowly in frequency~\cite{2019PhRvD.100l2002A}. The audio analogue of this is a tone that wanders in frequency; a note that changes pitch. Unlike the above Section~\ref{sec:single_tone} though, finding the signal isn’t as simple as searching for a peak in the spectrum for the full observing run.

One method (among many) for continuous wave searches is to split the observed signal into intervals of time, compute the Fourier transform for each time interval, and then create a grid (known as a spectrogram) where each cell is the Fourier amplitude of a particular frequency at a particular time.
This method turns the problem into finding the best path through the spectrogram grid as if it were a weighted graph, where here the weights are the Fourier amplitudes.

When finding the path, the method requires a limit for the allowed frequency change over time, usually informed by a physical model. Here, this frequency wander limit was set to 0.11 Hz per second to make it simply one frequency bin per every time bin, which will prove poorly adjusted to the signal.

The Viterbi algorithm is a path-finding algorithm that is used in continuous wave searches~\cite{viterbi_application} to solve this problem.
Abstractly speaking, the algorithm is given a weighted graph (e.g.\ the spectrogram grid), a sensible sequence of subgraphs (e.g.\ the columns of the spectrum at each time), and restrictions on connectivity (e.g.\ the allowed amount frequency wander). Then, at each iteration, it will find the best path to each node in the next subgraph, all the way from the first subgraph. At the end, it selects the overall best path from the first to the last subgraph (here, from the start to end time), this overall best path is called the Viterbi path. See Appendix~\ref{app:viterbi} for the full details of the algorithm in this particular implementation.

@@

We demonstrate this method of using the Viterbi algorithm by injecting a wandering frequency signal into the speaker and recording the output via the webcam.
Note that when creating a wandering frequency signal, the phase change must be integrated up over time, as detailed in Appendix~\ref{app:phase_gotcha}.

\begin{figure}[h]
	\includegraphics[width=\textwidth]{figures/expt_overlay_2_viterbi_test_webcam.pdf}
	\caption{Spectrogram of observed signal overlaid with pink dots for the injected signal and white crosses for the Viterbi path. Note the high correlation between them. Also note the deviations at the start time when the signal moves too quickly and just before the background anomaly at 150s.}
	\label{fig:viterbi_overlay}
\end{figure}

Shown in Figure~\ref{fig:viterbi_overlay} is the resultant spectrogram of the observed signal, overlaid with the injected signal (the pink dots) and the recovered Viterbi path (the white crosses).
To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment. Quantifying this, the Viterbi path stays within one frequency bin (approximately 0.3Hz) of the injected tone for 94\% of the run.
Note that the signal initially moves faster than the frequency wander limit, causing a discrepancy between the injected and recovered signals. To fix this, the limit should be doubled, although this would then cause the recovered path to stray more often. There also appears to be an anomaly at 150s, likely from a disturbance (such as someone walking) around the interferometer.

Empirically, decreasing the signal-to-noise ratio closer to unity (by making the speaker quieter) causes the Viterbi algorithm to stray more from the signal. However, since the only correlation through the grid is the signal, we always expect the Viterbi path to lie near to the signal no matter how high the noise is, and this is indeed what we see.

@@

The natural succession to observing simple tones is to study complex audio signals, such as music and speech.
The Viterbi algorithm and tracking more broadly is not applicable to these signals. Compare a single tone that slowly, continuously changes frequency to music and speech which consist of a broad spectrum that is constantly changing.
Rather than track them, the interest here is in being able to record and play-back these complex audio signals. Using the interferometer to do so makes it an ‘optical microphone’, using light to capture sound. This device has precedence in the laser microphones~\cite{laser_microphone} of the defence industry which operate on a variety of related principles.

Although this is indirectly related to the advanced signal processing required to extract gravitational wave signals, using an interferometer to capture sound can serve as an interesting demonstration to a broader physics and engineering audience. As such this could be introduced as a separate demonstration.

@@

For the optical microphone to be sensitive to speech and music it needs a frequency response (a sampling rate) far higher than the 30Hz of the webcam. In the context of audible frequencies (roughly 20Hz to 20kHz), speech intelligibility requires frequencies up to 3kHz and music requires frequencies up to and beyond 8kHz~\cite{speech_intelligibility}. Therefore the sampling rate of the optical microphone must be at least 16kHz to capture both speech and music. We achieved this with a photodiode.

A photodiode is an electrical component that acts as a regular diode, blocking any current flow in the `reverse' direction, when no light is incident on it. But as the intensity of incident light rises, it becomes increasingly conductive in the reverse direction. 
% Fundamentally, a current is created in the photodiode by the photo-electric effect.

We placed a (OSRAM BPW21) photodiode in reverse-bias over an (LM358) op-amp to create a photo-detector that would produce a voltage dependent on the incident intensity. We placed the photodiode in the interference pattern at roughly the same off-centre spot as the webcam was monitoring previously (although we saw no difference when monitoring at other positions). The photodiode was mounted on a cloth screen re-purposed from the dismantled commercial speaker, with the electrical leads connected underneath.

The voltage signal from the photo-detector was captured by a (MCP3008) 10-bit analog-to-digital converter (ADC) connected to a Raspberry Pi Model 3 v1.2. Together, these sampled the signal at roughly 16kHz, our desired limit.
